---
title: "Class11: AlphaFold2 Analysis"
author: "Job Rocha PID: A59023124 email: jrochahernandez@ucsd.edu"
format: pdf
---

Here we post process and inspect our modelling results from AlphaFold2 (AF).

My results from AF live in the folder/directory `HIV_dimer_23119`

```{r}
results_dir <- "HIV_dimer_23119/"

pdb.files <- list.files(results_dir, pattern=".pdb", full.names=TRUE)
```

We first need to align and supperpose these PDB models and we can use the `pdbaln()` function for this:

```{r}
library(bio3d)
pdbs <- pdbaln(pdb.files, fit=TRUE, exefile="msa")
```
## The RMSD matrix

A common measure of structural dis-similarity is called RMSD (root mean square distance).

```{r}
rd <- rmsd(pdbs)
rd
```

```{r}
#install.packages("pheatmap")
library(pheatmap)

rownames(rd) <- paste0("m", 1:5)
colnames(rd) <- paste0("m", 1:5)
pheatmap(rd)
```

Let's view these in Mol*. Here we want the filled coords.

```{r}
xyz <- pdbfit(pdbs, outpath="fitted")
```

A full atom based fitting or supperposition did not work very wll because we have multiple chains that are in different conformations.

I want to focus our supperposition on the most invariant part (the rigid "core" if you will).

```{r}
core <- core.find(pdbs)
```

```{r}
core.inds <- core
```

```{r}
xyz <- pdbfit(pdbs, inds=core.inds, outpath="core_fitted")
```

To evaluate how good multi-chain or multi-domain models are we need to look at the PAE scores (predicted aligned error).

There are output as JSON format files. Let's find all their file names:

```{r}
pae_files <- list.files(results_dir, pattern="0.json", full.names=TRUE)
```

```{r}
library(jsonlite)


pae1 <- read_json(pae_files[1], simplifyVector=TRUE)
pae5 <- read_json(pae_files[5], simplifyVector=TRUE)
```

```{r}
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae5$pae, xlab="Residue No.", ylab="Residue No.")
```

```{r}
plot.dmat(pae1$pae, xlab="Residue No.", ylab="Residue No.", zlim=c(0, 30))
```

# Main points

We can AlphaFold on google compute infrastructure :-)
We can read these results into R and process to help us make sense of these models and their PAE.